---
title: "cor2"
author: "Ben Holt"
date: "4/25/2020"
output: html_document
---
```{r}
library(RCurl)
library(ape)
sqTree<-read.tree(text=getURL("https://raw.githubusercontent.com/lukejharmon/pcm/master/datafiles/squamate.phy"))
plot(sqTree)
sqTree$tip.label
sqData<-read.csv(text=getURL("https://raw.githubusercontent.com/lukejharmon/pcm/master/datafiles/brandley_table.csv"))
```
Getting limbless traits...
```{r}
library(phytools)
limbless<-as.numeric(sqData[,"FLL"]==0 & sqData[,"HLL"]==0)
sum(limbless)
# get names that match
nn<-sqData[,1]
nn2<-sub(" ", "_", nn)
names(limbless)<-nn2

td<-treedata(sqTree, limbless)

plotSimmap(make.simmap(sqTree, limbless), pts=FALSE, fsize=0.8)
sqData
```


Determining ER rates
```{r}
library(geiger)
library(corHMM)
limt<-as.data.frame(td$data)
names(limt)
rownames(limt)<-td$phy$tip.label
limtdf<-data.frame(G_s=rownames(limt), limbless=limt$V1)
rate.mat.er<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ER")
```
What does this matrix mean?
## Number of changes needed to get from state 1 to state 2 and vice versa. ie: going from 1->2 requires one 'change' while 2->1 requires also one, since it is an equal rate model.
```{r}
lim.ER<-corHMM(td$phy,limtdf[, c(1,2)],rate.cat=1,rate.mat=rate.mat.er,node.states="marginal")
print(lim.ER)
```
What do these results mean?
## This is the maximum likeliood estimates of the transition rates from 0 (limbed)-> 1 and 1 (limbless)-> 0. The NA represent no change, as expected. Two rates are the same given the equal rate model parameter.  IE: gain and loss of limbs in equally likely..It should be noted that 'limbless' is defined in this context for 0 lenght lim measurement.

```{r}
rate.mat.ard<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ARD")
print(rate.mat.ard)

```
And these?
# The ARD model allows for 'all rates to differ', thus the transtion rate to 2 could be different than the transition rate to 1, while no changes remain NA. 

```{r}
lim.ARD<-corHMM(td$phy,limtdf[, c(1,2)],rate.cat=1,rate.mat=rate.mat.ard,node.states="marginal")
print(lim.ARD)
```
# transition rate is slightly higher for shits from limbed to limbless

which model is better?
```{r}
lim.ARD$AICc
lim.ARD$AIC
lim.ER$AICc
lim.ER$AIC
```
# Unlike the primate model, the ARD had a much lower AICc value for this example, indicating it is the better model. interesting! 

Now let’s look at multiple traits.

This is a matrix with four states
```{r}
rate.mat.er.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ER")
print(rate.mat.er.4state)
```

Convert the two binary traits into a single four character state. First I gotta generate another trait. For kicks and giggles, lets do body length (TL), though I know this is NEVER acceptable
```{r}
sqData
mean(sqData$TL) # 117.2395

short.tail<-as.numeric(sqData[,"TL"]<117.2395)
sum(short.tail) #165
mean(sqData$SVL)
short.body<-as.numeric(sqData[, 'SVL']<145.8368)
dats<-cbind(limbless, short.tail, short.body)

td<-treedata(sqTree, dats)# okay....
td$data
dim(td$data)
squamdat<-as.data.frame(td$data)
squamdat$Genus_spec<-rownames(squamdat)
rownames(squamdat)<-NULL
?data.frame
squamdat<-data.frame(Genus_species=squamdat$Genus_spec, limbs=squamdat$limbless, body=squamdat$short.body)
```

Now for models
```{r}
fourstate.trait<-rep(NA,Ntip(td$phy))
for(i in sequence(Ntip(td$phy))) {
    if(squamdat[i,2]==0 && squamdat[i,3]==0) {
        fourstate.trait[i]<-0
    }
    if(squamdat[i,2]==0 && squamdat[i,3]==1) {
        fourstate.trait[i]<-1
    }
    if(squamdat[i,2]==1 && squamdat[i,3]==0) {
        fourstate.trait[i]<-2
    }
    if(squamdat[i,2]==1 && squamdat[i,3]==1) {
        fourstate.trait[i]<-3
    }
}
fourstate.data<-data.frame(Genus_sp=squamdat$Genus_species, T1=fourstate.trait)

print(rayDISC(td$phy, fourstate.data, ntraits=1, model="ER", node.states="marginal"))
print(rayDISC(td$phy, fourstate.data, ntraits=1, rate.mat=rate.mat.er.4state, node.states="marginal", model="ARD"))

rate.mat.ard.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ARD")
print(rate.mat.ard.4state)
```


Now let’s make the equivalent of a GTR matrix:
```{r}
rate.mat.gtr.4state<-rate.mat.ard.4state
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(1,4))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(2,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(3,8))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(4,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(5,7))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(6,7))
print(rate.mat.gtr.4state)

print(rayDISC(td$phy, fourstate.data, ntraits=1, rate.mat= rate.mat.gtr.4state, node.states="marginal", model="ARD"))
```

Now make a model like Pagel 1994
```{r}
print(rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=2, nstates=2, model="ARD"))
rate.mat.pag94<-rate.par.drop(rate.mat.ard.4state, drop.par=c(3,5,8,10))
print(rate.mat.pag94)
```

Now that you have some introduction, there are two routes:

##Route 1

Construct a model to test if state 1 can never be lost

Experiment with the effects of frequencies at the root.

Create and use a model to see if transitions from 00 go to 11 only via 01.

##Route 2

Maddison and FitzJohn (2015) pretty convincingly show (to me) that Pagel (1994) is just not a good method. Ok. So work on a fix. They point to Read and Nee (1995) as a low power but possible solution. Look at their appendix, especially, and write an implementation.

