---
title: "Discrete Trees"
output: html_document
---
Let's say were interested in direct vs larval development in Desmognathine salamanders.
```{r}
library(geiger)
## getting a tree for Desmognathus salamanders. Testing the evolution of larval vs direct development
GetTreeFromOpenTree <- function(species) {
    library(ape)
    library(rotl)
  spec.id <- tnrs_match_names(species)$ott_id 
  spec.tree <- tol_subtree(ott_id=spec.id)}
tree<-GetTreeFromOpenTree('Desmognathus')
tree
plot.phylo(tree)
tree<-ape::compute.brlen(tree)
labs<-tree$tip.label
write.csv(labs, file = 'des_labs.csv')
```
Now add the discrete trait in as a column and add it back in
```{r}
discrete<-read.csv(file="/Users/benjaminholt/Desktop/Spring2020/Phylogenetics/des_dev.csv", stringsAsFactors = F)
```

select nodes to extract and trim the discrete data accordingly
```{r}
plot.phylo(tree)
tree2<-extract.clade(tree, root.edge = 0, collapse.singles = TRUE,interactive = TRUE)
plot.phylo(tree2)
CleanData <- function(phy, data) {
  plot.phylo(phy)
  data2<-data
  data2<-data2[, 2:ncol(data2)]
  rownames(data2)<-data[, 1]
  tree.dat<-treedata(tree2, data2, sort = T)
  return(tree.dat)
}
cleaned.tree<-CleanData(phy=tree, data = discrete)
tree<-cleaned.tree$phy
dat<-cleaned.tree$data
```

Make sure the traits of interest pop up where they should...

```{r}
library(phytools)
VisualizeData <- function(phy, data) {
  blah<-as.factor(setNames(data[,1],rownames(data)))
  dots<-dotTree(phy, blah,colors=setNames(c("blue","green"),levels(blah)), ftype="i",fsize=0.7)
}

VisualizeData(phy=tree, data =dat) ## this will give you an error, but still plots what I need
```


```{r}
library(phangorn)
plot.phylo(tree)
dat<-dat[,1]
levels(dat)<-c('larval', 'direct')
cleaned.discrete.phyDat <- phangorn::phyDat(dat, type="USER", levels = levels(dat)) #phyDat is a data format used by phangorn
anc.p <- phangorn::ancestral.pars(tree, cleaned.discrete.phyDat)
plotAnc(tree, anc.p, 1)
```

Not a great example because of the trait i used...
```{r}
anc.ml <- ancestral.pml(pml(tree, cleaned.discrete.phyDat), type="ml")
summary(anc.ml)

plotAnc(tree, anc.ml, 1)
```

How does this differ from parsimony?

# Parsimony applies the framework that the simplest explanation for taxonomic observation is most correct. This can be misleading from ancestral state because a single evolutionary event that causes changes early in the clade does not always represent the ancestral state. 

Why does it differ from parsimony?
# it relies heavily on the trait observed at the tips and less on the likelihood of the branches. For example, if 80% of a clade have trait A, parsimony would say trait A is ancestral, however with likelihood we can parse out and show that trait A is most commonly due to a change that had large effects, followed by diversification within that clade.

What does uncertainty mean?
Now, to the biological questions. For many of these, corHMM will be a useful package. Do the following analyses:

How can you estimate transition rates between states? Do it.
How could you examine if transition rates are equal? 
Think about the Lewis (2001) MKV model. Are your traits all variable? Will using this make sense for your data? Try using it. Do results change?

```{r}
library(corHMM)
dat<-cleaned.tree$data
dat<-as.data.frame(dat)
lh<-dat$life_history
names(lh)
mod1<-fitDiscrete(tree, lh, model = 'ER')
lh2<-as.numeric(lh)
is.rooted(tree)
treeu<-unroot(tree)
is.rooted(treeu)
mod2<-lewisMkv(treeu, lh2)
print(mod1)
```


How could you test order of state evolution?

####How can you estimate transition rates between states ? By using the fitdiscrete function in geiger(## we can do that by fitting a  model with different backward & forward rates

```{r}

ard.mod<-fitDiscrete(cleaned.tree$phy, cleaned.tree$data[,1], model = 'ARD')
print(ard.mod)

````

